The form of uncertainty affects selection for social learning

### 2022-06-21 ms.tex
Updates from Overleaf

Social learning is essential to survival. It is likely to evolve when it is more
efficient than asocial, trial-and-error learning. Theoretically, social learning
is adaptive for some uncertainty, but too much uncertainty makes
social information unreliable. This fact is empirically
supported across biology and human sciences. However, it is unclear how
specific classes and types of uncertainty affect social
learning evolution. Furthermore, existing models and experimental operationalizations of
uncertainty are ambiguously related, and 
tend to only consider a small number of behavioral
choices.  Here we use evolutionary agent-based modeling to consolidate
models of uncertainty in social learning evolution to address these issues.
We model societies of agents who perform one of a number
of possible behaviors to acquire payoffs in a time-varying environment.
We identified complex patterns in social learning evolution depending on contextual
uncertainty variables: environmental variability, ambiguous payoffs, 
decision sets size, and effective lifespans. 
Our work advances social learning evolution theory in a way that could help 
guide human, non-human, and artificial intelligence towards optimal responses 
to existential threats and new opportunities.\footnote{This document contains
\total{citnum} references.}  
\end{abstract}

@@

\cm{Uncertainty -> social learning widespread belief. But variously operationalized or not real uncertainty. Selling points of this paper 1) real uncertainty, 2) 4 kinds uncertainty, 3) cognition (soft-max temperature varies how much explores).}

Social learning enhances problem solving when acquiring information from others is more efficient than learning on one's own~\cite{Laland2004}. However, social learning can also lead individuals astray if, for example, they are copying outdated information. Theoretical models have helped clarify the circumstances under which social learning should be evolutionarily favored. For example, environments that change across time favor learning from others (as opposed to having a genetically fixed behavior) ~\cite{BoydRicherson1985,Henrich1998}, so long as they are not so variable that the social information reflects outdated and maladaptive behavior

point~\cite{Rogers1988,Feldman1996}. \cm{check that the citations still accurate}. This theoretical work has motivated much empirical research and helped predict behavior in several species ~\cite{McElreath2005,Kendal2018,Allen2019}.

From this literature, the idea that \emph{uncertainty} supports social learning has proliferated, but the term often conflates several features of the environment, including temporal variability,
spatial heterogeneity, ambiguity in behavioral payoff structure, and difficulty of the task.  In translating mathematical models to verbal predictions, the formal meaning of terms like \emph{uncertainty} is often lost, facilitating such conflation. Furthermore, as the number of formal models of social learning have expanded, an ever increasing number
of modeling choices ~\cite[Figure 1]{Kendal2018} and formalizations of \emph{uncertainty} has made it difficult to compare across models and consolidate our understanding of the contexts in which why social learning is adaptive.

@@

\mt{This paragraph is a first try to introduce the four uncertainty dimensions
with examples to illustrate the problem we are answering.
I outlined the next paragraph to get into more details of social learning:
how cultural evolution works; conformity, success-biased, etc. Cristina, please
edit this.} 
To systematically study features of uncertainty and its consequences for social learning, we identified four common classes of ecological uncertainty (``principal components'' of uncertainty) used in evolutionary models: (1) temporal environmental variability; (2) payoff
ambiguity (difference between best behavior choice and others); (3) 
selection set size (number of ecologically possible behaviors); and 
(4) the lifespan of agents (number of individual learning opportunities).
These were chosen because modeling and empirical 
studies across taxa tend use one or more of these uncertainty dimensions [Figure 1].

\cm{Other choices: softmax: existing models underestimate the power of learning from experience can improve both individual and helps social learners recover from outdated / misleading social info , no conformism: but that's ok because we're giving social learning the best chance of success 
no horizontal transmission, no spatial variation}

Temporal environmental variability is usually modeled as an independent probability of the environment changing its state each generation ~\cite{Feldman1996,McElreath2005}. The probabilities of environmental change are usually fixed ~\cite{BoydRicherson1985, Kendal2009}, but they have also been modeled as deterministic cycles ~\cite{Feldman1996} or as variable rates \cm{who does this?}. Furthermore, the consequences of environmental change can range from mild to  catastrophic. In the latter case a change of environment corresponds to no one having the adaptive behavior for the environment ~\cite{Rogers1988}.

or may be
modeled as occurring with a fixed or variable
probability---in these examples environments
can only be in one of two states. Payoff ambiguity sometimes takes the form
of differences in expected payoffs arising from different
behaviors or strategies~\cite{Enquist2007,Rendell2010}
or by varying the standard deviation of payoffs~\cite{McElreath2005}. 
Selection set size often takes only one value: 
for example, in empirical studies of \emph{bombus terresteris} (bumble
bee)~\cite{Baracchi2018} and \emph{parus major} (great tit)~\cite{Aplin2017},
experimenters provided two behaviors for
the bees or birds to perform, respectively, each yielding greater or smaller
payoffs depending on experimental treatment. Human studies have used two
or three possible behaviors~\cite{McElreath2005,Toyokawa2019}.
Modeling studies have used two 
behaviors~\cite{Feldman1996,Rendell2010} and four behaviors~\cite{Enquist2007}. 
The age or lifespan of individuals was
found to affect on social learning in both \emph{parus
major}~\cite{Aplin2017} and \emph{poecilia reticulata} 
(guppies)~\cite{Leris2016} \mt{Need better connection between lifespan and 
age, or better references to support use of lifespan}.
Shorter lifespans mean individuals
die more uncertain of which behaviors are optimal, and pass that uncertainty on
to their offspring. To answer our research questions about the combined effect
of various forms of uncertainty on social learning evolution, we developed an
agent-based model that incorporates all four of
these uncertainty variables. We systematically vary these parameters
to understand and predict their effects on social learning evolution.

\paragraph{Social learning evolution} (CRISTINA)
To add to confusion about how to integrate diverse models of social learning
evolution, different models select different social learning components in
seemingly \emph{ad hoc} ways 
\begin{itemize}
  \item 
    Vertical, oblique, horizontal transmission
  \item
    Conformity, payoff-biased, frequency-dependent, etc. A note about how
    ``conformity'' is often not distinguished from other forms of social learning
  \item
    Review human studies
  \item
    Review non-human studies~\cite{Leris2016,Aplin2017,Avargues-Weber2018,Baracchi2018}
  \item
    One sentence on how dual-inheritance theory enables us to gloss over whether
    social learning is genetically or culturally evolved.
  \item
    More things from which we selected the operation of our model
\end{itemize}

@@

@@

To answer our research questions we developed an agent-based model of social
learning evolution in which a society of $N$ individuals each must decide which of
$B$ behaviors to perform at each time step within a generation of $L$ time steps. At
the end of each generation, agents are selected with replacement from the population
to reproduce another $N$ agents. Agents who inherit the social learning trait then
learn about behavioral payoffs from the previous generation, while asocial learners
begin life with no \emph{a priori} knowledge of the world.  Each behavior is a
``bandit'', a common modeling and experimental approach for representing behaviors
with probabilistic
payoffs~\cite{SuttonBartoBook,McElreath2005,Steyvers2009,Rendell2010,Schulz2019}.  Each behavior
$b$ yields Bernoulli payoffs: payoff of 1 with probability $\pi_b$ and zero payoff
with probability $1 - \pi_b$; $\pi_b$ is therefore also the expected payoff of
behavior $b$. One behavior in each generation is optimal and pays off $\pihigh$,
while the rest pay off $\pilow < \pihigh$.  Model agents must decide which behavior
to perform at each time step.  To do this, agents use an explore-exploit strategy to
sometimes try the most profitable behavior they know about, and other times try
alternatives that may pay off more reliably.  We developed a series of computational
analyses by systematically varying uncertainty parameters and observing the
frequency that model populations evolve to be social learners, the average payoffs
accrued by agents in each setting, and the time it takes for model populations to
reach fixation (agents all social or all asocial learners). Model dynamics are
illustrated in Figure~\ref{fig:schematic}.

>>>>

Social learning is essential to human survival. It is likely to evolve when it is more
efficient than asocial, trial-and-error learning. Theoretically, social learning
is adaptive for some uncertainty, but too much uncertainty makes
social information unreliable. This fact is empirically
supported across biology and human sciences. However, it is unclear how
specific classes and types of uncertainty affect social
learning evolution. Furthermore, existing models and experimental operationalizations of
uncertainty are ambiguously related, and 
tend to only consider a small number of behavioral
choices.  Here we use evolutionary agent-based modeling to consolidate
models of uncertainty in social learning evolution to address these issues.
We model societies of agents who perform one of a number
of possible behaviors to acquire payoffs in a time-varying environment.
We identified complex patterns in social learning evolution depending on contextual
uncertainty variables: environmental variability, ambiguous payoffs, 
decision sets size, and effective lifespans. 
Our work advances social learning evolution theory in a way that could help 
guide human, non-human, and artificial intelligence towards optimal responses 
to existential threats and new opportunities.\footnote{This document contains
\total{citnum} references.}  
\end{abstract}

@@

\cm{Notes on our June 2 discussion (USPs): "Uncertainty favors social learning" is a widespread belief. But it is operationalized in different ways, and oftentimes it does not reflect REAL uncertainty. Selling points of this paper 1) real uncertainty, 2) 4 kinds uncertainty, 3) cognition (soft-max temperature varies how much explores).}

Social learning enhances problem solving when acquiring information from others is more efficient than learning on one's own~\cite{Laland2004}. However, social learning can also lead individuals astray if, for example, they are copying outdated information. Theoretical models have helped clarify the circumstances under which social learning should be evolutionarily favored. For example, environments that change across time favor learning from others (as opposed to having a genetically fixed behavior) ~\cite{BoydRicherson1985,Henrich1998}, so long as they are not so variable that the social information reflects outdated and maladaptive behavior point~\cite{Rogers1988,Feldman1996}. \cm{check that the citations still accurate}. This theoretical work has motivated much empirical research and helped predict behavior in several species ~\cite{McElreath2005,Kendal2018,Allen2019}.

From this literature, the idea that \emph{uncertainty} supports social learning has proliferated, but the term often conflates several features of the environment, including temporal variability,
spatial heterogeneity, ambiguity in behavioral payoff structure, and difficulty of learning the task.  In translating mathematical models to verbal predictions, the formal meaning of terms like \emph{uncertainty} is often lost, facilitating such conflation. Furthermore, as the number of formal models of social learning have expanded, an ever increasing number
of modeling choices ~\cite[Figure 1]{Kendal2018} and formalizations of \emph{uncertainty} has made it difficult to compare across models and consolidate our understanding of the contexts in which social learning is adaptive.

@@

\mt{This paragraph is a first try to introduce the four uncertainty dimensions
with examples to illustrate the problem we are answering.
I outlined the next paragraph to get into more details of social learning:
how cultural evolution works; conformity, success-biased, etc. Cristina, please
edit this.} 
To systematically study features of uncertainty and its consequences for social learning, we identified four common classes of ecological uncertainty used in evolutionary models: (1) temporal environmental variability; (2) selection set size (number of possible behaviors); (3) payoff
ambiguity (difference in the reward between adopting the best behavior and alternatives); and (4) the lifespan of agents (number of individual learning opportunities).
These were chosen because modeling and empirical 
studies across taxa tend use one or more of these uncertainty dimensions (Figure \ref{fig:schematic}).

Temporal environmental variability is fundamental to most models of social learning, yet even this is implemented in several ways. This is usually modeled as an independent probability of the environment changing its state each generation ~\cite{Feldman1996,McElreath2005}. The probabilities of environmental change are usually fixed ~\cite{BoydRicherson1985, Kendal2009}, but they have also been modeled as deterministic cycles ~\cite{Feldman1996} or as variable rates \cm{who does this?}. Furthermore, the consequences of environmental change can range from mild to  catastrophic. In the latter case a change of environment results in no one having the adaptive behavior for the environment ~\cite{Rogers1988}. Such catastrophic environmental changes present a large adaptive challenge as individuals cannot bank on accumulated information from previous generations. Other models of environmental change introduce less uncertainty as they fluctuate between two environmental states with corresponding adaptive behaviors. This means that all individuals with previously maladaptive behaviors then have adaptive behaviors upon the environment changing ~\cite{Perreault2012}.

Selection set size often takes only one value: 
for example, in empirical studies of \emph{bombus terresteris} (bumble
bee)~\cite{Baracchi2018} and \emph{parus major} (great tit)~\cite{Aplin2017},
experimenters provided two behaviors for
the bees or birds to perform, respectively, each yielding greater or smaller
payoffs depending on experimental treatment. Human studies have used two
or three possible behaviors~\cite{McElreath2005,Toyokawa2019}.
Modeling studies have used two 
behaviors~\cite{Feldman1996,Rendell2010} and four behaviors~\cite{Enquist2007}.

Payoff ambiguity differs from selection set size in that it not only affects uncertainty, it also directly influences the strength of selection. This parameter sometimes takes the form
of differences in expected payoffs arising from different
behaviors or strategies~\cite{Enquist2007,Rendell2010}
or by varying the standard deviation of payoffs~\cite{McElreath2005}.

The age or lifespan of individuals was
found to affect on social learning in both \emph{parus
major}~\cite{Aplin2017} and \emph{poecilia reticulata} 
(guppies)~\cite{Leris2016} \mt{Need better connection between lifespan and 
age, or better references to support use of lifespan}.
Shorter lifespans mean individuals
die more uncertain of which behaviors are optimal, and pass that uncertainty on
to their offspring. To answer our research questions about the combined effect
of various forms of uncertainty on social learning evolution, we developed an
agent-based model that incorporates all four of
these uncertainty variables. We systematically vary these parameters
to understand and predict their effects on social learning evolution.

\paragraph{Social learning evolution} (CRISTINA) 
\cm{I think a lot of this may be too much detail. I'm still editing and focusing the above on the main forms of uncertainty we are modeling and reviewing the relevant literature as we go. Yes we've made choices about things like horizontal vs vertical, but i think we can just justify that when we introduce the model }
To add to confusion about how to integrate diverse models of social learning
evolution, different models select different social learning components in
seemingly \emph{ad hoc} ways 
\begin{itemize}
  \item 
    Vertical, oblique, horizontal transmission
  \item
    Conformity, payoff-biased, frequency-dependent, etc. A note about how
    ``conformity'' is often not distinguished from other forms of social learning
  \item
    Review human studies
  \item
    Review non-human studies~\cite{Leris2016,Aplin2017,Avargues-Weber2018,Baracchi2018}
  \item
    One sentence on how dual-inheritance theory enables us to gloss over whether
    social learning is genetically or culturally evolved.
  \item
    More things from which we selected the operation of our model
\end{itemize}

@@

\cm{Other choices we made that might need to be justified: 1) softmax: existing models underestimate the power of learning from experience by not having a rich cognition. This rich cognition can improve both individual learning and help social learners recover from outdated / misleading social info. 2) no conformism: but that's ok because we're giving social learning the best chance of success. 3) no horizontal transmission, no spatial variation}

@@

To answer our research questions we developed an agent-based model of social
learning evolution in which a society of $N$ individuals \cm{do we ever define N used?} each must decide which of
$B$ behaviors to perform at each time step within a generation of $L$ time steps. At
the end of each generation, agents are selected with replacement from the population
to reproduce another $N$ agents. Agents who inherit the social learning trait then
learn about behavioral payoffs from the previous generation, while asocial learners
begin life with no \emph{a priori} knowledge of the world.  Each behavior is a
``bandit'', a common modeling and experimental approach for representing behaviors
with probabilistic
payoffs~\cite{SuttonBartoBook,McElreath2005,Steyvers2009,Rendell2010,Schulz2019}.  Each behavior
$b$ yields Bernoulli payoffs: payoff of 1 with probability $\pi_b$ and zero payoff
with probability $1 - \pi_b$; $\pi_b$ is therefore also the expected payoff of
behavior $b$. One behavior in each generation is optimal and pays off $\pihigh$,
while the rest pay off $\pilow < \pihigh$.  Model agents must decide which behavior
to perform at each time step.  To do this, agents use an explore-exploit strategy to
sometimes try the most profitable behavior they know about, and other times try
alternatives that may pay off more reliably.  We developed a series of computational
analyses by systematically varying uncertainty parameters and observing the
frequency that model populations evolve to be social learners, the average payoffs
accrued by agents in each setting, and the time it takes for model populations to
reach fixation (agents all social or all asocial learners). Model dynamics are
illustrated in Figure~\ref{fig:schematic}.
